Causal Discovery on Precarity and Depression

### 2025-02-11 draft_v1/draft_v1.tex
update_results
only FCI in main text + proportion matrix plots

\date{2025-02-08}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precariousness and
depression is critical for developing effective interventions. This
study utilizes data from the HELIUS cohort study to explore these
relationships using advanced causal discovery methods. By applying
algorithms such as FCI, and CCI, and combining traditional Gaussian CI
tests with non-parametric approaches like RCoT, we investigate how
different indicators of precariousness---including factors related to
employment, social relations and relational stress, financial situation,
and housing---affect depressed mood, both as a sum score and at the
individual symptom level. Our findings reveal that relational stress
consistently emerges as a potential causal factor for depressed mood,
while symptoms such as sleep disturbances, guilt, and anhedonia are
particularly sensitive to external stressors, acting as potential early
warning signals or intervention points for prevention. Moreover, the
results highlight complexities in the data, including the influence of
latent confounders and the challenges of capturing cyclic relationships.
Despite some limitations, such as unresolved ambiguities in causal
directions and challenges with mixed data distributions, this study
demonstrates the utility of causal discovery tools in disentangling the
intricate interplay between social and mental health dynamics. By
mapping these causal structures into computational models, future
research can simulate intervention effects, providing actionable
insights to mitigate the impact of precariousness on mental health. This
study serves as a foundational effort, offering both methodological
advancements and practical implications for addressing depression at a
population level.
\end{abstract}

@@

Mental health problems in urban areas have been reported to be on the
rise the complexity of mental health systems presents significant
challenges in understanding the underlying mechanisms driving these
issues, let alone planning effective interventions.

Research has aimed to identify underlying factors contributing to mental
health problems. Recent research looked into the association between
mental health and indicators of precariousness in various dimensions of
life, indicating a high level of uncertainty and instability in people's
lives, such as factors related to employment, social relations,
finances, housing, and culture (ref Leonie's paper). This comprehensive
perspective on precariousness helped to highlight how different aspects
of life may be interconnected. While this research has advanced our
understanding of which indicators of precariousness are related to
mental health, a key question remains unanswered: how do these factors
influence mental health? Specifically, the lack of directional
information --- knowing what influences what --- limits our ability to
identify and prioritize effective intervention targets.

This study aims to investigate the causal relationships between
different dimensions of precariousness and mental health outcomes, to
with a specific focus on depression, the most prevalent mental health
issue. Using causal discovery methods, we explore how different aspects
of precariousness influence depression and delve deeper into the
dynamics at the symptom level. By examining individual depressive
symptoms, we aim to identify which symptoms may act as initiators by
being particularly sensitive to precariousness. Through this analysis,
our goal is to uncover the causal mechanisms underlying mental health
challenges and provide a foundation for developing more effective and
targeted interventions.

@@

There are numerous causal discovery algorithms available; however, in
this study, we focus on algorithms suited to the potential cyclic
relationships within our system. Specifically, we use FCI (Fast Causal
Inference) and CCI (Cyclic Causal Inference), both capable of accounting
for such cycles under certain conditions (Mooij \& Claassen, 2020;
Strobl, 2019). Additionally, we include the PC algorithm as a reference,
given its simplicity and prominence as one of the most widely known
causal discovery methods (Spirtes et al., 2001).

@@

As shown in Table 1, the resulting graphs from FCI and CCI differ
slightly (\emph{PAG}: partial ancestral graph; \emph{MAAG}: maximal
almost ancestral graph) due to their reliance on different underlying
assumptions. Despite these differences, both graphs are forms of
\emph{ancestral graphs}, designed to encode causal relationships between
variables, where the presence of an edge indicates causal
\emph{ancestry.} Directed edges, \(A \stararrow B\), indicate that \(B\)
is not an ancestor of \(A\) in every graph within the Markov equivalence
class, \(Equiv(G)\). The Markov equivalence class refers to the set of
graphs that encode the same conditional independence relationships, such
that the same \emph{d-separation} relations hold across all graphs in
the class (Spirtes et al., 2001). \(A \startail B\) indicates that \(B\)
is an ancestor of \(A\) across all graphs in \(Equiv(G)\). Circle
endpoints, \(A \starcirc B\), represent ambiguity in the ancestral
relationship, meaning \(B\)'s ancestral status relative to \(A\) varies
across graphs in \(Equiv(G)\).\footnote{\(*\) serves as a
  \textit{meta-symbol}, representing one of the three possible
  edge-endpoints. For instance, \(A \tailstar B\) can indicate any of
  the following edges: \(A\) \textemdash~\(B\), \(A\tailarrow B\), or
  \(A \tailcirc B\) (Park et al., 2024).} When \(A \arrowarrow B\),
neither \(A\) nor \(B\) is an ancestor of the other, indicating the
presence of a latent confounder between them.

The algorithms also differ in their approaches and assumptions for
detecting cycles. In CCI's MAAG, \(A\) \textemdash~\(B\) indicates that
\(A\) is an ancestor of \(B\) and \(B\) is an ancestor of \(A\),
referring to a cycle between \(A \stackedarrows B\). FCI, on the other
hand, identifies potential cycles more subtly: fully-connected nodes
with circle endpoints (\(\circirc\)) in FCI's PAG may suggest a cyclic
structure. FCI basically provides a sufficient condition to distinguish
nodes that are not part of a cycle, offering a more nuanced approach
(Mooij \& Claassen, 2020). Table 1 highlights further differences in the
assumptions underlying these algorithms. CCI operates under the
assumption of a linear system, while FCI, particularly for inferring
cyclic relationships, assumes a non-linear system without selection bias
and adheres to more general faithfulness and Markov conditions (i.e.,
\emph{\(\sigma\)-separation} and \emph{\(\sigma\)-faithfulness} setting)
(Forré \& Mooij, 2018). When the respective assumptions of each
algorithm are satisfied, their outcomes for cyclic relationships align
with those illustrated in Figure~\ref{fig-examplegraphs}. While an
exhaustive discussion of causal discovery concepts and
algorithm-specific details is beyond the scope of this paper, readers
seeking a deeper understanding are encouraged to consult Park et al.
(2024) for comprehensive insights into these methods and their
applications.

@@

The graph produced by the PC algorithm is a \emph{CPDAG} (completed
partially directed acyclic graph), where directed edges (\(A\) → \(B\))
indicate that \(A\) is a direct cause (parent) of \(B\). Unlike FCI and
CCI, the CPDAG does not include circle symbols. Instead, when the PC
algorithm cannot determine the direction of causality, it represents
this uncertainty with bidirectional arrows. While the PC algorithm
serves as a useful reference, its strict assumptions of acyclicity and
the absence of latent confounders restrict its applicability in more
complex scenarios. Therefore, our primary focus remains on the results
from FCI and CCI, with all PC algorithm results provided in the
\hyperref[sec-appendix]{Appendix} for completeness.

One practical challenge in applying these algorithms to the HELIUS
dataset is that the data does not follow a Gaussian distribution, and
the relationships among variables are unlikely to be strictly linear. To
account for this, we supplement the commonly used Gaussian conditional
independence (CI) test, which relies on partial correlations, with a
non-parametric CI test based on reproducing kernel methods (Zhang et
al., 2012). While kernel-based conditional independence tests (KCIT) are
effective, they are computationally intensive for large datasets like
HELIUS, as their quadratic scaling with sample size is primarily driven
by the inversion of large kernel matrices. To mitigate this issue, we
use the Randomized Conditional Correlation Test (RCoT), which
approximate kernel methods using random Fourier features. This approach
reduces computation to linear scaling with sample size, significantly
lowering computational costs (Strobl et al., 2019). For a more detailed
explanation of RCoT, refer to Section~\ref{sec-rcot}.

\subsection{Analysis}\label{analysis}

We analyze the causal structure using three approaches: examining the
relationship (1) between the five individual precarity factors and the
PHQ sum score, which represents overall depression severity; (2) between
individual symptoms and precarity factors; and (3) between individual
symptoms and overall precarity (the sum of the five precarity factors).
Using the PHQ sum score simplifies the analysis by reducing
dimensionality, offering computational efficiency and a broad,
interpretable perspective on the relationship between depression and
precarity factors. In contrast, analyzing individual symptoms provides a
more detailed understanding by capturing the diverse ways symptoms
respond to different precarity factors. However, this detailed approach
introduces methodological challenges, including non-standard
distributions of individual symptom variables and the complexities
associated with increased dimensionality. Finally, examining individual
symptoms in relation to the overall precarity score complements the
analysis by revealing how symptoms collectively connect to general
precarity. Aggregating precarity factors may uncover relationships that
could be overlooked in analyses of individual factors, particularly when
a symptom has associations across multiple factors that are not strong
enough to emerge in individual analyses. By integrating these three
approaches, we achieve a balance between capturing detailed granularity
and gaining a broader understanding of the relationship between
precarity and depression.

Furthermore, to address sensitivity to parameter values and enhance the
robustness of our results, we implement multiple settings combined with
bootstrapping. For each of the 100 bootstrap samples, we estimate causal
graphs and retain only the edges and directions that exceed predefined
thresholds. The analyses are conducted under the following conditions:

@@

This setup yields 16 combinations (2 significance levels × 4 thresholds
× 2 CI tests), applied across three algorithms. Each combination is
repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per
algorithm. For analyses involving individual symptom variables, we
simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing
the number of bootstrap samples to 30, easing computational demands. To
further enhance efficiency, we fix the skeleton of edges among symptom
variables based on the common structure estimated across all algorithms.
This approach minimizes spurious edges in the symptom network and
significantly reduces the computational time needed for skeleton
estimation. This fixed structure also aligns well with commonly reported
skeleton structures in the literature, ensuring consistency with
existing findings (\textbf{cite the comp\_model paper}).\footnote{We
  fixed the skeleton of the symptom network using a common structure
  derived from the PC, FCI, and CCI algorithms with an alpha level of
  0.001. This approach preserved the complexity of symptom interactions,
  enabling unrestricted estimation of causal directions both among
  symptoms (\emph{symptom --- symptom} interactions) and between
  symptoms and precarity factors.} Lastly, to summarize the results, we
identify the most frequently occurring edge endpoints across different
experimental setups.\footnote{The detailed proportion of each edge
  endpoint occurrence is provided in \textbf{?@sec-propmatrix}.} This
ensures that only stable and consistent edges are retained, providing a
clearer and more reliable understanding of the relationships between
precarity factors and depression. See Figure~\ref{fig-workflow} for an
overview of the analysis workflow.

@@

The sum score graphs provide a high-level summary of how precarity
factors collectively influence overall depression severity, focusing on
aggregated relationships. Figure~\ref{fig-sum} illustrates the causal
relationships between precarity factors (\emph{P.hou}, \emph{P.emp},
\emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the depression sum score
(\emph{PHQsum}) under two different setups: (a) using both Gaussian CI
test and RCoT, and (b) using RCoT alone. Black edges represent
consistent causal relationships identified by both FCI and CCI, while
gray dashed edges denote inconsistent relationships that vary between
the two methods. The inconsistent endpoints of the gray edges are marked
with circles.

In graph (a), the key pathways suggest that employment precarity
(\emph{P.emp}) and social precarity (\emph{P.soc}) do not cause
depression (\emph{PHQsum}). \emph{P.soc} does not cause recent
relational stress (\emph{S.rel}) or financial stress (\emph{S.fin}), and
these stressors are likely related by a latent confounder. Additionally,
\emph{P.emp} and \emph{S.fin} are identified as non-causes of housing
precarity (\emph{P.hou}). Both FCI and CCI detect a dependency between
\emph{P.emp} and \emph{S.fin}, but they disagree on the direction of the
relationship, leaving it unclear whether \emph{S.fin} causes
\emph{P.emp} or whether a latent confounder mediates their relationship.
Similar ambiguities are found in the connections between \emph{S.rel}
and \emph{PHQsum} and between \emph{S.fin} and \emph{PHQsum}.

Graph (b), derived solely from the nonparametric RCoT method. Both
graphs consistently identify that \emph{P.emp} is not a cause of
\emph{P.hou} or \emph{PHQsum}, and that \emph{P.soc} does not cause
\emph{PHQsum} or \emph{S.fin}. The relationship between stressors and
depression remains unresolved between the two algorithms. However, the
graph derived from RCoT provides greater confidence that depression is
not the cause of \emph{S.fin} and suggests a stronger likelihood that
the financial stress may contribute to depression or that their
relationship is mediated by a latent confounder. The role of relational
stress has become less pronounced, as the edge between \emph{P.soc} and
\emph{S.rel} is omitted, and the relationship between \emph{S.rel} and
\emph{S.fin} is now inconsistent between the algorithms.

Overall, both graphs together highlight the potential roles of stressors
like \emph{S.fin} and \emph{S.rel}, which are closely tied to
depression, either causally or via latent confounders. Employment and
social precarity may also be influenced by depression, either directly
or through latent variables. Lastly, housing precarity does not directly
impact depression but remains connected to employment precarity and
financial stress, likely mediated by direct or latent pathways.

@@

Moving from the sum score representation to the symptom-level graph
provides a more granular perspective on the causal relationships between
precarity factors and depression. This approach highlights the
heterogeneity in how precarity factors influence individual depressive
symptoms --- \emph{slp} (sleep), \emph{ene} (energy), \emph{app}
(appetite), \emph{mot} (motor), \emph{sui} (suicidal), \emph{anh}
(anhedonia), \emph{glt} (guilt), and \emph{dep} (depressed mood). While
the sum score graph aggregates all symptoms into a single
measure---potentially obscuring nuanced relationships---the symptom
graph uncovers distinct pathways for different symptoms. In the
symptom-level graph (Figure~\ref{fig-sym}), consistent relationships are
represented by black solid edges, while areas of disagreement between
the FCI and CCI algorithms are denoted by gray dashed edges. Endpoints
marked with circles indicate differences in directional conclusions
between the algorithms. Additionally, the navy dashed edges represent
relationships unique to the graphs generated using Gaussian CI testing.

The symptom-level graph reveals a complex and interconnected structure,
far more intricate than the sum score graph. Its denser network
highlights the strong interdependence among symptoms and suggests the
presence of latent confounding influences, as indicated by numerous
bidirectional arrows.

Certain nodes in the network emerge as more \emph{causally} central,
highlighting their potential importance as intervention points. Among
the depressive symptoms, \emph{anh}, \emph{dep}, \emph{slp}, and
\emph{glt} emerge as particularly influential, with multiple outgoing
edges (o-\textgreater) to other symptoms, suggesting their roles as
potential key drivers within the symptom network. Especially, symptoms
such as \emph{glt}, \emph{slp}, and \emph{anh} are particularly
critical, potentially acting as initiator or activator nodes within the
network due to their apparent connections with precarity factors. These
symptoms appear to be especially sensitive to external stressors,
potentially manifesting early in response to such conditions and
subsequently activating other interconnected symptom nodes.

The causal structure involving precarity factors in the symptom-level
graph is largely consistent with the patterns observed in the sum score
graphs. As in the sum graphs, relational stress (\emph{S.rel}) emerges
as a potential causal factor for depression, while employment
(\emph{P.emp}) and social precarity (\emph{P.soc}) are more likely to be
influenced by depressive symptoms. The symptom-level analysis, however,
provides more specificity by pinpointing the symptoms involved in these
relationships. For instance, \emph{slp} is identified as influencing
\emph{P.emp} or potentially through a latent confounder (\emph{slp}
o-\textgreater{} \emph{P.emp}) , while \emph{glt} appears to affect
\emph{P.soc} or may be linked via a confounder (\emph{glt}
o-\textgreater{} \emph{P.soc}). Additional edges are observed in the
graph generated using Gaussian CI tests, such as \emph{slp}
\textless-\textgreater{} \emph{S.rel} and \emph{slp} o-\textgreater{}
\emph{P.emp}. As seen in the sum graph, Gaussian CI testing tends to
produce a denser graph. In this case, causal relationships involving
\emph{slp} are particularly prominent.

Some discrepancies, however, exist between the symptom-level and sum
graphs. For example, financial stress (\emph{S.fin}) does not have any
edges with depressive symptoms in the symptom-level graph, although it
maintains associations with other precarity factors. Additionally,
housing precarity (\emph{P.hou}) becomes entirely disconnected from the
rest of the network, appearing as an isolated node.

>>>>

\date{2025-02-11}
\begin{document}
\maketitle
\begin{abstract}
\noindent Understanding the causal mechanisms linking precariousness and
depression is critical for developing effective interventions. This
study utilizes data from the HELIUS cohort study to explore these
relationships using advanced causal discovery methods. By applying
algorithms such as FCI, and CCI, and combining traditional Gaussian CI
tests with non-parametric approaches like RCoT, we investigate how
different indicators of precariousness---including factors related to
employment, social relations and relational stress, financial situation,
and housing---affect depressed mood, both as a sum score and at the
individual symptom level. Our findings reveal that relational stress
consistently emerges as a potential causal factor for depressed mood,
while symptoms such as sleep disturbances, guilt, and anhedonia are
particularly sensitive to external stressors, acting as potential early
warning signals or intervention points for prevention. Moreover, the
results highlight complexities in the data, including the influence of
latent confounders and the challenges of capturing cyclic relationships.
Despite some limitations, such as unresolved ambiguities in causal
directions and challenges with mixed data distributions, this study
demonstrates the utility of causal discovery tools in disentangling the
intricate interplay between social and mental health dynamics. By
mapping these causal structures into computational models, future
research can simulate intervention effects, providing actionable
insights to mitigate the impact of precariousness on mental health. This
study serves as a foundational effort, offering both methodological
advancements and practical implications for addressing depression at a
population level.
\end{abstract}

@@

Mental health problems in urban areas are on the rise, yet the
complexity of mental health systems poses significant challenges in
understanding the underlying mechanisms driving these issues --- let
alone designing effective interventions. Research has long sought to
identify underlying factors contributing to mental health problems.
Recent research looked into the association between mental health and
indicators of precariousness in various dimensions of life, indicating a
high level of uncertainty and instability in people's lives, such as
factors related to employment, social relations, finances, housing, and
culture (ref Leonie's paper). This comprehensive perspective on
precariousness helped to highlight how different aspects of life may be
interconnected. While this research has advanced our understanding of
which indicators of precariousness are related to mental health, a key
question remains unanswered: how do these factors influence mental
health? Specifically, the lack of directional insight --- knowing what
influences what --- limits our ability to identify and prioritize
effective intervention targets.

This study aims to investigate the causal relationships between
different dimensions of precariousness and mental health outcomes, to
with a specific focus on depression, the most prevalent mental health
issue. Using causal discovery methods, we explore how various aspects of
precariousness contribute to depression and whether depression, in turn,
exacerbates precariousness, potentially creating a feedback loop.
Furthermore, we zoom in at the symptom level, examining individual
depressive symptoms to identify those that may act as initiators ---
triggering the activation of the symptom network --- by displaying
heightened sensitivity to specific precariousness factors. We also
investigate symptoms that may function as bridging elements, linking the
dynamics of depression with precariousness by feeding back into
precarious conditions, potentially reinforcing the cycle of mental
health deterioration and precarious conditions. Through this analysis,
we aim to uncover the causal mechanisms driving mental health challenges
related to precariousness and establish a foundation for more targeted
and effective interventions by identifying the key driving forces within
this system.

@@

There are numerous causal discovery algorithms available; however, in
this study, we focus on algorithms suited to the potential cyclic
relationships within our system. Specifically, we use \emph{FCI} (Fast
Causal Inference) and \emph{CCI} (Cyclic Causal Inference), both capable
of accounting for such cycles (Mooij \& Claassen, 2020; Strobl, 2019).
Additionally, we include the \emph{PC} algorithm as a reference, given
its simplicity and prominence as one of the most widely known causal
discovery methods (Spirtes et al., 2001).

@@

As shown in Table 1, the resulting graphs from FCI and CCI differ
slightly (\emph{PAG}: partial ancestral graph; \emph{MAAG}: maximal
almost ancestral graph) due to their reliance on different underlying
assumptions. Despite these differences, both graphs belong to the class
of \emph{ancestral graphs}, which are designed to encode causal
relationships between variables, where the presence of an edge indicates
causal \emph{ancestry.} In these graphs, directed edges,
\(A \stararrow B\), indicate that \(B\) is not an ancestor of \(A\) in
every graph within the Markov equivalence class, \(Equiv(G)\). The
Markov equivalence class represents a set of graphs that encode the same
conditional independence relationships, ensuring that the same
\emph{d-separation} conditions hold across all graphs in the class
(Spirtes et al., 2001). Conversely, an edge marked as \(A \startail B\)
indicates that \(B\) is an ancestor of \(A\) across all graphs in
\(Equiv(G)\). Circle endpoints, \(A \starcirc B\), represent ambiguity
in the ancestral relationship, meaning \(B\)'s ancestral status relative
to \(A\) varies across graphs in \(Equiv(G)\).\footnote{\(*\) serves as
  a \textit{meta-symbol}, representing one of the three possible
  edge-endpoints. For instance, \(A \tailstar B\) can indicate any of
  the following edges: \(A\) \textemdash~\(B\), \(A\tailarrow B\), or
  \(A \tailcirc B\) (Park et al., 2024).} Finally, when an edge is
represented as \(A \arrowarrow B\), it implies that neither \(A\) nor
\(B\) is an ancestor of the other, suggesting the presence of a latent
confounder influencing both variables.

The algorithms also differ in how they detect cycles and represent
cyclic relationships in their resulting graphs. In CCI's MAAG, \(A\)
\textemdash~\(B\) indicates that \(A\) is an ancestor of \(B\), and
simultaneously \(B\) is an ancestor of \(A\), referring to a cyclic
relationship between \(A \stackedarrows B\). FCI, on the other hand,
identifies potential cycles more subtly. In its PAG, fully-connected
nodes with circle endpoints (\(\circirc\)) may suggest the presence of
cyclic structures. FCI, therefore, provides a sufficient condition to
distinguish variables that are not part of a cycle, offering a more
nuanced approach to handling cyclic relationships (Mooij \& Claassen,
2020).

Table 1 highlights further differences in the assumptions underlying
these algorithms. CCI operates under the assumption of a linear system,
whereas FCI, particularly when used to infer cyclic relationships,
assumes a non-linear system without selection bias and adheres to more
general faithfulness and Markov conditions (i.e.,
\emph{\(\sigma\)-separation} and \emph{\(\sigma\)-faithfulness} setting)
(Forré \& Mooij, 2018). When the respective assumptions of each
algorithm are met, their inferred cyclic relationships align with those
illustrated in Figure~\ref{fig-examplegraphs}.

While CCI's MAAG provides an explicit representation of cyclic
relationships, it has certain theoretical limitations, as the MAAG may
not always fully preserve d-separation relations from the original
graph, \(\mathcal{G}\) (Strobl, 2019). Given these considerations, we
examine and compare results from both algorithms, placing more weight on
the FCI results, which are presented in the main results section, while
discussing CCI results where relevant, with detailed findings provided
in the \hyperref[sec-appendix]{Appendix}. A full discussion of causal
discovery concepts and algorithmic details is beyond the scope of this
paper; however, readers seeking a more in-depth understanding can refer
to Park et al. (2024) for a comprehensive exploration of these methods
and their applications.

@@

The graph produced by the PC algorithm is a \emph{CPDAG} (completed
partially directed acyclic graph), where directed edges (\(A\) → \(B\))
indicate that \(A\) is a direct cause (parent) of \(B\). Unlike FCI and
CCI, the CPDAG does not include circle symbols. Instead, when the PC
algorithm cannot determine the direction of causality, it represents
this uncertainty with bidirectional arrows. While the PC algorithm
serves as a useful reference, its strict assumptions --- acyclicity and
the absence of latent confounders---limit its applicability in more
complex scenarios. For this reason, our primary focus remains on the
results obtained from FCI and CCI, with all PC algorithm results
provided in the \hyperref[sec-appendix]{Appendix} for completeness.

One practical challenge in applying these algorithms to the HELIUS
dataset is that most variables do not follow a Gaussian distribution,
and the relationships among variables are unlikely to be strictly
linear. To address this, we supplement the commonly used Gaussian
conditional independence (CI) test, which relies on partial
correlations, with a non-parametric CI test based on reproducing kernel
methods (Zhang et al., 2012). While kernel-based conditional
independence tests (KCIT) are theoretically effective in capturing
non-linear dependencies and subtle interactions in data with
non-Gaussian noise, they are computationally demanding for large
datasets like HELIUS. Their quadratic scaling with sample size is
primarily due to the need to invert large kernel matrices, making them
impractical for large-scale analysis (Rahimi \& Recht, 2007). To
mitigate this issue, we use the Randomized Conditional Correlation Test
(RCoT), which approximates kernel methods using random Fourier features.
This approach reduces computational complexity from quadratic to linear
scaling with sample size, significantly lowering computational costs
while maintaining robustness (Strobl et al., 2019). For a more detailed
explanation of RCoT, refer to Section~\ref{sec-rcot}.

\subsection{Analysis}\label{sec-analysis}

We analyze the causal structure using three complementary approaches:
(1) Examining the relationship between the five individual
precariousness factors and the PHQ sum score, which represents overall
depression severity. (2) Investigating the relationship between
individual depressive symptoms and individual precariousness factors.
(3) Assessing the relationship between individual depressive symptoms
and overall precariousness (the sum of all precariousness factors).
Using the PHQ sum score as a measure of depression simplifies the
analysis by reducing dimensionality, improving computational efficiency,
and providing a broad, interpretable perspective on how precariousness
factors relate to overall depression severity. However, this aggregated
approach may overlook granular symptom-level relationships.

By zooming in on individual symptoms, we capture a more detailed picture
of how specific depressive symptoms interact with different
precariousness factors. This approach allows us to map individual causal
pathways, identifying which symptoms are particularly sensitive to
certain precarious conditions and vice versa. However, this level of
detail introduces methodological challenges, including non-standard
distributions of symptom variables and the high-dimensional nature of
the analysis, which can reduce statistical power and significantly
increase computational demands---especially when using non-parametric
conditional independence (CI) tests.

Finally, analyzing individual symptoms in relation to overall
precariousness provides an integrative perspective, revealing how
depression symptoms collectively relate to precarious conditions. This
approach complements the fully disaggregated analysis by aggregating
precariousness factors, potentially uncovering relationships that may be
too weak to detect when examined separately. This is particularly
relevant when a symptom is associated with multiple precariousness
factors, but these associations are not strong enough to emerge in an
individual-level analysis. By integrating these three approaches ---
ranging from partially aggregated to fully disaggregated --- we aim to
achieve a balance between capturing detailed granularity and gaining a
broader understanding of the relationship between precarity and
depression.

Furthermore, to address sensitivity to parameter values and enhance the
robustness of our results, we implement multiple settings combined with
bootstrapping. For each condition, we generate 100 bootstrap samples,
estimate causal graphs, and retain only the edge-endpoints that exceed
predefined thresholds. If no edge-endpoint type surpasses the threshold,
we assign a circle (\(\circ\)) to indicate uncertainty in the edge
direction.

The analyses are conducted under the following conditions:

@@

This setup yields 16 combinations (2 significance levels × 4 thresholds
× 2 CI tests), applied across three algorithms. Each combination is
repeated for 100 bootstrap samples, yielding a total of 1,600 graphs per
algorithm. For analyses involving individual symptom variables, we
simplify the setup by focusing on thresholds of 0.6 and 0.7 and reducing
the number of bootstrap samples to 30, easing computational demands. To
further enhance efficiency, we fix the skeleton of the symptom network
using a common structure derived from PC, FCI, and CCI algorithms
(\(\alpha\) = 0.01, RCoT CI test). This step constrains which edges can
be present among symptoms but does not restrict the estimation of causal
directions within the symptom network. Additionally, the entire causal
structure between symptoms and precariousness factors remains fully
estimable, ensuring that key relationships are still inferred. Beyond
its computational benefits --- particularly in reducing the time
required for skeleton estimation --- this fixed structure also aligns
well with commonly reported symptom network structures in the literature
(Park et al., 2025), ensuring consistency with existing findings.

To obtain a final, stable summary of causal relationships, we identify
the most frequently occurring edge-endpoints across different
experimental setups. First, within each condition, we determine the most
frequent edge-endpoint from all bootstrap samples. Then, across
different conditions, we aggregate these selected symbols and determine
the overall most common edge-endpoint for each graph entry.

Since we have an even number of experimental setups, ties may
occasionally occur. In such cases, we explicitly denote them in the
resulting graph using dashed lines. To further help with the
interpretation and understanding of the resulting graph, we provide a
corresponding table for each graph, showing the proportion of each
edge-endpoint type to clarify how dominant a given symbol is in each
case.

By double summarizing the results in this way, we generate one final
graph per algorithm for each analysis approach: the partially aggregated
analysis, which uses either the sum of precariousness factors or the sum
of depression factors to capture overall trends, and the fully
disaggregated analysis, which examines individual precariousness and
depression factors separately. This approach ensures that only stable
and consistent edges --- those that persist across different parameter
settings --- are retained, leading to a robust and reliable
understanding of the relationships between precariousness factors and
depression. See Figure~\ref{fig-workflow} for an overview of the
analysis workflow.

@@

The sum score graphs provide a high-level summary of how precarity
factors collectively influence overall depression severity, focusing on
aggregated relationships. Figure~\ref{fig-sum} illustrates the causal
relationships between precariousness factors (\emph{P.hou},
\emph{P.emp}, \emph{P.soc}, \emph{S.rel}, \emph{S.fin}) and the
depression sum score (\emph{PHQsum}), as identified by the FCI
algorithm. As described in Section~\ref{sec-analysis}, dashed edges
indicate ties, with the exact edge endpoint where the tie occurs is
represented by a bold horizontal bar.

The causal graph in panel (a) highlights key pathways in the
relationships between precarity factors and depression (PHQsum).
Employment precarity (P.emp) and social precarity (P.soc) are not
identified as causes of depression, whereas financial stress (S.fin)
appears to play a causal role. While P.emp and P.soc are generally not
recognized as causes of other precariousness factors, S.fin emerges as a
potential cause, as indicated by its circle edge endpoint. Supporting
this, the proportion matrix plot in panel (b) shows that S.fin has some
probability of causally influencing either P.emp or P.soc.

The matrix plot can be interpreted such that the symbol in
matrix{[}\(i\), \(j\){]} represents the relationship
\(i -~[symbol] \  j\). For example, if matrix{[}\(i\), \(j\){]} is
\textgreater{} and matrix{[}j, i{]} is \(\circ\), then the inferred
relationship is \(i \circarrow j\). In the table, different edge types
are represented by distinct colors: light gray for the absence of an
edge, blue for circles, green for arrowheads, and coral for arrowtails.
The colors are blended based on the proportion of each edge type, with
higher proportions increasing the opacity of the corresponding color,
making dominant symbols more visually prominent.

Relational stress (S.rel) plays a more nuanced role, interacting with
depression and S.fin through a latent confounder. Its relationship with
P.soc remains less certain, as the proportion table suggests a fair
probability that this connection is absent, uncertain, or bidirectional.
Meanwhile, housing precarity (P.hou) is not directly causally related to
depression but is linked to P.emp and S.fin. While P.emp and S.fin are
identified as non-causes of P.hou, it remains unclear whether P.hou
causally influences P.emp or if their relationship is mediated by an
unobserved confounder.

The findings from the CCI algorithm (see Figure~\ref{fig-sum-cci}) are
largely consistent with those from the FCI algorithm, with one
exception: CCI does not identify S.fin as a cause of depression.
Additionally, CCI introduces greater uncertainty in edge directions,
with more ties mainly between circle and arrowhead endpoints, and tends
to favor arrowheads more frequently than FCI. Despite these differences,
the skeleton structure and overall causal directions derived from CCI
align well with the results of the FCI algorithm, supporting the key
pathways.

@@

Moving from the sum score representation to the disaggregated
symptom-level graph provides a more granular perspective on the causal
relationships between precariousness factors and depression symptoms.
Unlike the sum score graph, which aggregates all symptoms into a single
measure---potentially obscuring nuanced relationships---the
symptom-level graph highlights heterogeneity in how precariousness
factors influence individual depressive symptoms: con (concentration),
slp (sleep), ene (energy), app (appetite), mot (motor), sui (suicidal),
anh (anhedonia), glt (guilt), and dep (depressed mood), and vice versa.

The symptom-level graph in Figure~\ref{fig-sym} panel (a) reveals a more
complex and interconnected structure than the sum score graph,
reflecting the strong interdependence among symptoms and suggesting much
presence of latent confounders, as indicated by numerous bidirectional
edges. As before, dashed edges denote ties, with specific ties marked by
bold horizontal bars. In Figure~\ref{fig-sym} panel (b), the proportion
matrix provides more insight into these directionalities. Here, opaque
green dominates most symptom-to-symptom connections, indicating that
arrowheads are the most frequently inferred edge type. However, blue
regions, which correspond to circle endpoints, are particularly common
for anh, slp, ene, and sui, reflecting uncertainty in the causal
direction among these symptoms. Notably, the only arrowtail connection
appears between dep and anh, suggesting that anhedonia (anh) is a
potential cause of depressed mood (dep).

Looking at the symptom-precariousness connections, one of the key
patterns is that ties are most frequently found in the relationships
between individual symptoms and precariousness factors. This suggests
that these causal links may be less stable across different conditions.
Upon closer examination, most ties arise due to discrepancies between
the Gaussian CI test and the RCoT test---with Gaussian CI favoring
arrowheads and RCoT more often predicting the absence of an edge.

This discrepancy likely arises from a combination of factors. First, as
the analysis shifts from aggregated to disaggregated variables, the
number of conditional independence (CI) tests increases, leading to a
loss of statistical power in high-dimensional causal discovery. With
more variables being conditioned on, the data becomes sparser, making it
harder to detect weaker dependencies. Additionally,
over-conditioning---controlling for too many variables---can
artificially remove statistical associations, leading to false
negatives. Beyond this general statistical power issue, the fundamental
differences in how these CI tests operate contribute to the
inconsistency. RCoT, as a nonparametric test, does not assume linearity
and can capture both linear and nonlinear dependencies. However, it is
more conservative and requires larger sample sizes, making it more prone
to false negatives in high-dimensional settings. In contrast, partial
correlation (Gaussian CI test) assumes linear Gaussian relationships,
making it more permissive and sometimes detecting weak dependencies that
may not be statistically meaningful. As a result, when Gaussian CI
detects an edge while RCoT does not, it could indicate that the
relationship is weak, strictly linear, or requires more data for
reliable nonparametric detection. This explains why many ties in the
proportion matrix show a 50-50 split between symbols---where half of the
conditions with RCoT suggest the absence of an edge, while the other
half using Gaussian CI test favor an arrowhead. This systematic
divergence highlights the challenges of causal discovery in
high-dimensional settings, where varying statistical assumptions can
lead to differing conclusions about the presence and direction of causal
relationships.

Despite these differences, some consistent patterns emerge across both
CI tests, particularly in the case of financial stress (S.fin), which
appears to be connected to nearly all other variables---though many of
these connections are marked by ties, reflecting uncertainty in
directionality. One exception is the stronger evidence suggesting that
S.fin may cause changes in app (appetite), while its relationships with
other symptoms, such as ene, mot, and dep, remain ambiguous. Similarly,
social precarity (P.soc) exhibits numerous connections with depressive
symptoms, with most edges pointing toward P.soc rather than outward from
it. This suggests that depressive symptoms, particularly dep and glt,
may contribute to worsening social precarity rather than the other way
around. Additionally, anh also shows some probability of causally
influencing P.soc, reinforcing the idea that social precarity is more
often a consequence rather than a driver of depressive symptoms.

Other precariousness factors exhibit more uncertain but still notable
relationships. Relational stress (S.rel) shows weak but existing
connections with slp, glt, and app, though directionality remains
unclear in many cases. Employment precarity (P.emp) also connects to
symptoms like slp, mot, and con, but, like S.rel, these relationships
exhibit a 50/50 split in directionality, reflecting uncertainty in the
inferred causal paths. In line with the aggregated graph in
Figure~\ref{fig-sum}, housing precarity (P.hou) does not appear to have
any direct relationship with depression symptoms but consistently shows
associations with employment precarity (P.emp) and financial stress
(S.fin). The causal relationships among precariousness factors remain
largely unchanged from the aggregated analysis, with financial stress
(S.fin) exhibiting the strongest tendency to influence other
precariousness factors.
